---
title: "Another_File"
format: html
editor: visual
---

# MinneMUDAC 2025: Big Brother Big Sister Twin Cities

MinneMUDAC 2025 has teamed up with Big Brother Big Sisters Twin Cities for this year’s challenge.  Big Brother Big Sisters Twin Cities the largest and oldest youth award-winning mentoring organization in the greater Twin Cities.  Each year, we match up youth (Littles age 8-13) and their families with caring adults (Bigs) who volunteer as mentors.  Through a variety of community-based, school-based, and workplace-based mentoring programs, and together with our community, we want every youth to have a mentor, be affirmed in who they are, and explore who they want to be.

## Novice Division

#### Responsive Variables of Interest:

-   `Match Length`

-   `Closure Reason`

#### Libraries:

```{r}
library(tidyverse)
library(readxl)
library(naniar)
library(FSA)
```

#### Loading in the data:

```{r}
novice_data <- read_excel("../data/Novice.xlsx")
```

#### Quick Overview of The Data:

```{r}
glimpse(novice_data)
```

```{r}
summary(novice_data)
```

```{r}
head(novice_data)
```

#### Initial Findings:

**Rows:** 3,275

**Columns:** 66

**Key columns of interest:**

-   `Stage` – Whether the match is `Closed` or `Active`.

-   `Closure Reason` – Reason why the match ended.

-   `Match Activation Date` & `Match Closure Meeting Date` – Used to calculate match duration.

-   `Match Length` – Length of the match (in months).

-   `Big` and `Little` demographics and interest fields – Useful for understanding match characteristics.

    -   `Big Approved Date`, `Big Birthdate`, `Big Acceptance Date`, `Little Birthdate`, `Little Acceptance Date` – Useful for deriving age at match and match timelines.

    -   `Big Gender`, `Big Race/Ethnicity`, `Big Level of Education`, `Big County`, `Little Gender`, `Little Participant: Race/Ethnicity`, `Program` and `Program Type` – May influence match outcomes.

    -   `Big Age`, `Big Days Interview to Acceptance`, `Big Days Interview to Match`, `Match Length` – Numeric target for regression.

    -   `Big Re-Enroll` – Indicator variable.

**Missing Data:**

-   `Big Employer/School Census Block Group` and `Big Home Census Block Group` have significant missing values.

### Understand and analyze the response variable distributions:

#### Analyzing `Match Length` Distribution:

**From `summary()` above:**

-   `Match Length` ranges from 0 to 97.2 months.

-   Median: \~16.8 months

-   Mean: \~23.38 months (longer tail suggesting possible skew)

-   NA values: None

**Histogram**

```{r}
ggplot(novice_data, aes(x = `Match Length`)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Match Length",
       x = "Match Length (months)",
       y = "Frequency")
```

**Density Plot**

```{r}
ggplot(novice_data, aes(x = `Match Length`)) +
  geom_density(fill = "lightblue", alpha = 0.7) +
  labs(title = "Density of Match Length",
       x = "Match Length (months)",
       y = "Density")
```

#### Analyzing `Closure Reason` Distribution:

**From `summary()` above:**

-   `Closure Reason` is a categorical variable.

-   Unique values: 34

-   NA values: 786

**Bar Plot**

```{r}
novice_data |>
  filter(!is.na(`Closure Reason`)) %>%
  count(`Closure Reason`) %>%
  ggplot(aes(x = reorder(`Closure Reason`, -n), y = n, fill = `Closure Reason`)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Distribution of Closure Reasons",
       x = "Closure Reason",
       y = "Count") +
  theme(legend.position = "none")
```

Interesting how high the COVID closure reason is.

#### Double Checking `Match Length` and Missing Values:

Using `Match Activation Date` and `Match Closure Meeting Date` to calculate match duration.

```{r}
novice_data <- novice_data %>%
  mutate(Match_Length_Calculated = as.numeric(difftime(`Match Closure Meeting Date`, `Match Activation Date`, units = "days")) / 30.4)
```

**Plotting to Compare:**

```{r}
ggplot(novice_data, aes(x = `Match Length`, y = Match_Length_Calculated)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  labs(title = "Comparison: Provided vs Calculated Match Length",
       x = "Match Length (Provided)",
       y = "Match Length (Calculated)")
```

**Checking Missing Values:**

```{r}
sum(is.na(novice_data$`Match Length`)) 
```

```{r}
sum(is.na(novice_data$`Closure Reason`)) 
```

Using the `naniar` package to visualize missing values:

```{r}
vis_miss(novice_data, warn_large_data = FALSE)
```

Many columns contain plenty of missing values, enough that the data set as a whole is 55% missing.

### How do the response variable distributions vary across Program Type? 

#### `Match Length` Distribution by `Program Type`

**Box Plot for Match Length by Program Type**

```{r}
ggplot(novice_data, aes(x = `Program Type`, y = `Match Length`, fill = `Program Type`)) +
  geom_boxplot(outlier.color = "red", outlier.shape = 1, notch = TRUE) +
  labs(title = "Match Length Distribution by Program Type",
       x = "Program Type",
       y = "Match Length (months)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none")
```

**Violin Plot for Detailed Shape**

```{r}
ggplot(novice_data, aes(x = `Program Type`, y = `Match Length`, fill = `Program Type`)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  labs(title = "Density of Match Length by Program Type",
       x = "Program Type",
       y = "Match Length (months)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none")
```

#### `Closure Reason` Distribution by `Program Type`

**Stacked Bar Plot for Closure Reason**

```{r}
novice_data |>
  filter(!is.na(`Closure Reason`)) |>
  count(`Program Type`, `Closure Reason`) |>
  ggplot(aes(x = `Program Type`, y = n, fill = `Closure Reason`)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Closure Reason Distribution by Program Type",
       x = "Program Type",
       y = "Proportion",
       fill = "Closure Reason") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Faceted Bar Plot for Closure Reason Counts**

```{r}
novice_data |>
  filter(!is.na(`Closure Reason`)) |>
  count(`Program Type`, `Closure Reason`) |>
  ggplot(aes(x = `Closure Reason`, y = n, fill = `Closure Reason`)) +
  geom_bar(stat = "identity") +
  facet_wrap(~`Program Type`, scales = "free_y") +
  labs(title = "Count of Closure Reasons by Program Type",
       x = "Closure Reason",
       y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none")
```

#### Statistical Test for `Match Length` Variation

**ANOVA or Kruskal-Wallis Test**

-   ANOVA: If assumptions of normality and homogeneity of variance hold.

```{r}
anova_result <- aov(`Match Length` ~ `Program Type`, data = novice_data)
summary(anova_result)
```

-   **Exploration:**

    -   **F value**

        -   74.28 — High F-value means the difference between program types are much larger than within group variation.

    -   Pr(\>F)

        -   p-value \< 2e-16 — Very small p-value, meaning the differences between `Program Type` are highly significant.

        -   That means `Match Length` varies significantly by `Program Type`.

-   Kruskal-Wallis: If assumptions are violated (non-parametric alternative).

```{r}
kruskal_test <- kruskal.test(`Match Length` ~ `Program Type`, data = novice_data)
kruskal_test
```

-   **Exploration:**

    -   **Chi-squared value**

        -   182.07 — High value means strong difference in distributions.

    -   **p-value**

        -   p-value \< 2.2e-16 — Extremely small p-value.

            -   This confirms that there are statistically significant differences in `Match Length` between `Program Type` groups.

#### Statistical Test for `Closure Reason` Variation

Determining whether `Closure Reason` distribution varies significantly across `Program Type`.

```{r}
closure_table <- table(novice_data$`Closure Reason`, novice_data$`Program Type`)
chisq.test(closure_table)
```

-   **Exploration:**

    -   **Chi-squared value**

        -   The warning may mean that some expected counts may be too small, high chance for inaccuracies. Could try Fisher's Exact Test.

        -   1265.7 — Very high value means large difference between observed and expected frequencies.

    -   **p-value**

        -   p-value \< 2.2e-16 — Extremely small p-value.

            -   Shows that `Closure Reason` distributions are significantly different across `Program Type`.

#### Post-hoc Test/Analysis

For ANOVA we can run Tukey’s HSD to determine which `Program Types` differ significantly.

```{r}
# Resolving issue with column names having spaces causing an error.
colnames(novice_data) <- make.names(colnames(novice_data))
```

```{r}
anova_result <- aov(Match.Length ~ Program.Type, data = novice_data)
tukey_result <- TukeyHSD(anova_result)
print(tukey_result)
```

-   **Exploration:**

    -   **Significant Results (p \< 0.05):**

        -   `Site - Community`:

            -   Difference: -12.06

            -   95% CI: \[-14.34, -9.78\]

            -   p-value: **\< 0.001** — Highly significant, meaning that `Site` matches have shorter match lengths than `Community`.

        -   `Site Based Facilitated - Community`:

            -   Difference: -9.59

            -   95% CI: \[-12.67, -6.50\]

            -   p-value: **\< 0.001** — Significant, indicating that `Site Based Facilitated` matches have shorter match lengths than `Community`.

    <!-- -->

    -   **Non-Significant Results:**

        -   `Site Based Plus - Community`

            -   p = 0.92 (no significant difference)

        -   `Site Based Facilitated - Site`

            -   p = 0.28 (no significant difference)

        -   `Site Based Plus - Site`

            -   p = 0.31 (no significant difference)

        -   `Site Based Plus - Site Based Facilitated`

            -   p = 0.43 (no significant difference)

For Kruskal-Wallis we can use Dunn's Test for pairwise comparison.

```{r}
dunn_test <- dunnTest(`Match Length` ~ `Program Type`, data = novice_data, method = "bonferroni")
print(dunn_test)
```

-   **Exploration:**

    -   This test compares the groups to each other to identify how they differ.

    -   **Significant Comparisons:**

        -   `Community - Site`: p \< 0.001 — Highly significant difference.

        -   `Community - Site Based Facilitated`: p \< 0.001 — Significant.

        -   `Site - Site Based Facilitated`: p \< 0.001 — Significant.

    -    **Non-Significant Comparisons:**

        -   `Community - Site Based Plus`: p = 1.00 (no significant difference).

        -   `Site - Site Based Plus`: p = 0.60 (not significant).

        -   `Site Based Facilitated - Site Based Plus`: p = 1.00 (not significant).

    Community and Site matches have significantly different match lengths.

    Community and Site Based Facilitated are also different.

    Site and Site Based Facilitated differ too.

    But, Site Based Plus is not significantly different from other groups.

Trying out Fisher's Exact Test for a more accurate results due to the chi-squared warning.

```{r}
fisher_test <- fisher.test(closure_table)
fisher_test
```

The error above is because the data is too large for the Fisher test to handle the exact calculations. Trying again using simulated p-values.

```{r}
fisher_test <- fisher.test(closure_table, simulate.p.value = TRUE, B = 10000)
fisher_test
```

This shows that there is a significant relationship between the type of program and the reason for closure.

#### Plots to visualize findings:

```{r}
ggplot(as.data.frame(closure_table), aes(x = Var2, y = Freq, fill = Var1)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Program Type", y = "Count", fill = "Closure Reason") +
  theme_minimal()
```

```{r}
ggplot(novice_data, aes(x = Program.Type, y = Match.Length, fill = Program.Type)) +
  geom_boxplot() +
  labs(title = "Match Length Distributions Across Program Types", x = "Program Type", y = "Match Length (months)") +
  theme_minimal()
```

### Have the response variable distributions changed over time?

#### Analyzing `Match Length` Over Time:

Extracting year from `Match Activation Date` and `Match Closure Meeting Date`, gives us two approaches for `Match Length` over time.

```{r}
novice_data <- novice_data %>%
  mutate(Match_Activation_Year = year(as.Date(Match.Activation.Date)),
         Match_Closure_Year = year(as.Date(Match.Closure.Meeting.Date)))
```

Group by year and summarize mean match length

```{r}
match_length_trend_act <- novice_data %>%
  group_by(Match_Activation_Year) %>%
  summarise(
    mean_match_length = mean(Match.Length, na.rm = TRUE),
    median_match_length = median(Match.Length, na.rm = TRUE),
    count = n()
  )
```

```{r}
match_length_trend_clo <- novice_data %>%
  group_by(Match_Closure_Year) %>%
  summarise(
    mean_match_length = mean(Match.Length, na.rm = TRUE),
    median_match_length = median(Match.Length, na.rm = TRUE),
    count = n()
  )
```

Plotting Match Length trend over time

```{r}
ggplot(match_length_trend_act, aes(x = Match_Activation_Year, y = mean_match_length)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(
    title = "Average Match Length Over Time (Activation Year)",
    x = "Year",
    y = "Average Match Length (months)"
  ) +
  theme_minimal()
```

```{r}
ggplot(match_length_trend_clo, aes(x = Match_Closure_Year, y = mean_match_length)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(
    title = "Average Match Length Over Time (Closure Year)",
    x = "Year",
    y = "Average Match Length (months)"
  ) +
  theme_minimal()
```

#### Distribution of `Match Length` Across Years

```{r}
ggplot(novice_data, aes(x = as.factor(Match_Activation_Year), y = Match.Length)) +
  geom_boxplot(aes(fill = as.factor(Match_Activation_Year))) +
  labs(
    title = "Distribution of Match Length by Year (Activation Year)",
    x = "Year",
    y = "Match Length (months)",
    fill = "Year"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggplot(novice_data, aes(x = as.factor(Match_Closure_Year), y = Match.Length)) +
  geom_boxplot(aes(fill = as.factor(Match_Closure_Year))) +
  labs(
    title = "Distribution of Match Length by Year (Closure Year)",
    x = "Year",
    y = "Match Length (months)",
    fill = "Year"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Analyzing `Closure Reason` Changes Over Time

`Closure Reason` Trends (Activation Year)

```{r}
closure_trend_act <- novice_data %>%
  group_by(Match_Activation_Year, Closure.Reason) %>%
  summarise(count = n(), .groups = "drop")
```

`Closure Reason` Trends (Closure Year)

```{r}
closure_trend_clo <- novice_data %>%
  group_by(Match_Closure_Year, Closure.Reason) %>%
  summarise(count = n(), .groups = "drop")
```

Plots:

```{r}
ggplot(closure_trend_act, aes(x = Match_Activation_Year, y = count, fill = Closure.Reason)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Closure Reasons Distribution Over Time (Activation Year)",
    x = "Year",
    y = "Count",
    fill = "Closure Reason"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggplot(closure_trend_clo, aes(x = Match_Closure_Year, y = count, fill = Closure.Reason)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Closure Reasons Distribution Over Time (Closure Year)",
    x = "Year",
    y = "Count",
    fill = "Closure Reason"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
