---
title: "Undergrad_Division"
format: html
---

# MinneMudac 2025 Undergrad Division

### Initial Work:

**Loading Libraries:**

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(tidytext)
library(textdata)
library(sentimentr)
library(tm)
library(caret)
library(pROC)
library(SnowballC)
library(xgboost)
library(randomForest)
```

**Loading in Datasets:**

```{r warning=FALSE}
training_file <- "../data/Training.xlsx"
test_file <- "../data/Test_Truncated.xlsx"

train_data <- read_excel(training_file, sheet = 1)
test_data <- read_excel(test_file, sheet = 1)
```

```{r}
glimpse(training_data)
```

```{r}
glimpse(test_data)
```

**Checking for Missing Values:**

```{r}
colSums(is.na(training_data))
```

```{r}
colSums(is.na(test_data))
```

**NAs Observations:**

-   **Contact Notes:**

    -   `Match Support Contact Notes` has **1,454 missing values** in `training_data` and **100 missing values** in `test_data`.

    -   This could impact text analysis and sentiment analysis.

    -   We’ll handle missing notes by either:

    ```         
    -    Imputing with placeholder text (`"No Notes Available"`) or

    -    Dropping matches with missing notes, depending on the approach.
    ```

-   **Completion Date:**

```         
-    `Completion Date` has **147 NAs** in `training_data` and **1 NA** in `test_data`.

-    Since `Match Length` is derived from this, NAs here could impact the predictive model. We’ll likely **filter out these rows**.
```

-   **Big and Little Attributes:**

```         
-    Several columns like `Big Occupation`, `Big Car Access`, `Little Mailing Address`, etc., have high percentages of missing values (many over 50%).

-    These columns may need to be **dropped** or **imputed** depending on their correlation with the target variable (`Match Length`).
```

### Droping/Imputing NAs:

**Contact Notes:**

```{r}
training_data$`Match Support Contact Notes`[is.na(training_data$`Match Support Contact Notes`)] <- "No Notes Available"
test_data$`Match Support Contact Notes`[is.na(test_data$`Match Support Contact Notes`)] <- "No Notes Available"
```

**Completion Date:**

```{r}
training_data <- training_data %>%
  filter(!is.na(`Completion Date`))

test_data <- test_data %>%
  filter(!is.na(`Completion Date`))
```

**Dropping Columns with 50% or more missing values (saving just in case):**

```{r}
threshold <- 0.5 * nrow(training_data)

training_data <- training_data %>%
  select_if(~sum(is.na(.)) < threshold)

test_data <- test_data %>%
  select_if(~sum(is.na(.)) < threshold)
```

**Making a `Match Length` column:**

```{r}
training_data <- training_data %>%
  mutate(match_length = as.numeric(difftime(`Completion Date`, `Match Activation Date`, units = "days")))

test_data <- test_data %>%
  mutate(match_length = as.numeric(difftime(`Completion Date`, `Match Activation Date`, units = "days")))
```

**Calculate days between calls:**

```{r}
training_data <- training_data %>%
  group_by(`Match ID 18Char`) %>%
  arrange(`Completion Date`) %>%
  mutate(days_between_calls = as.numeric(difftime(`Completion Date`, lag(`Completion Date`), units = "days")))

training_data$days_between_calls[is.na(training_data$days_between_calls)] <- 0

test_data <- test_data %>%
  group_by(`Match ID 18Char`) %>%
  arrange(`Completion Date`) %>%
  mutate(days_between_calls = as.numeric(difftime(`Completion Date`, lag(`Completion Date`), units = "days")))

test_data$days_between_calls[is.na(test_data$days_between_calls)] <- 0
```

### Visualizing Days Between Calls:

```{r}
ggplot(training_data, aes(x = days_between_calls)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  theme_minimal() +
  labs(title = "Distribution of Days Between Calls", x = "Days", y = "Count")
```

```{r}
correlation_cadence <- cor(training_data$days_between_calls, training_data$match_length, use = "complete.obs")
print(paste("Correlation between days between calls and match length:", correlation_cadence))
```

Correlation: **0.583**, which indicates a **moderate to strong positive correlation**. The correlation is **not extremely high**, other factors likely influence match length

### Analyze Match Support Contact Notes:

```{r}
training_notes <- training_data %>%
  unnest_tokens(word, `Match Support Contact Notes`) %>%
  anti_join(stop_words, by = "word") %>%
  mutate(word = wordStem(word))
```

**Term Frequency and TF-IDF:**

```{r}
word_freq <- training_notes %>%
  count(word, sort = TRUE)
```

```{r}
tf_idf <- training_notes %>%
  count(`Match ID 18Char`, word, sort = TRUE) %>%
  bind_tf_idf(word, `Match ID 18Char`, n) %>%
  arrange(desc(tf_idf))

head(tf_idf, 20)
```

**Load sentiment lexicon:**

```{r}
bing_lexicon <- get_sentiments("bing")
```

**Perform sentiment analysis:**

```{r}
training_notes <- training_notes %>%
  distinct(`Match ID 18Char`, word, .keep_all = TRUE)
```

```{r}
sentiment_notes <- training_notes %>%
  inner_join(bing_lexicon, by = "word") %>%  # No more many-to-many issues!
  count(`Match ID 18Char`, sentiment) %>%
  spread(sentiment, n, fill = 0)
```

```{r}
training_notes <- training_notes %>%
  count(`Match ID 18Char`, word, name = "word_count")
```

```{r}
sentiment_notes <- training_notes %>%
  inner_join(bing_lexicon, by = "word") %>%
  count(`Match ID 18Char`, sentiment) %>%
  spread(sentiment, n, fill = 0)
```

```{r}
bing_lexicon %>%
  group_by(word) %>%
  filter(n() > 1) %>%
  arrange(word)
```

```{r}
bing_lexicon <- bing_lexicon %>%
  mutate(word = wordStem(word)) %>%
  distinct(word, .keep_all = TRUE)
```

```{r}
sentiment_notes <- training_notes %>%
  inner_join(bing_lexicon, by = "word") %>%
  count(`Match ID 18Char`, sentiment) %>%
  spread(sentiment, n, fill = 0)
```

**Joining data into training:**

```{r}
training_data <- training_data %>%
  left_join(sentiment_notes, by = `Match ID 18Char`)
```

```{r}
colnames(training_data) <- make.names(colnames(training_data), unique = TRUE)
colnames(test_data) <- make.names(colnames(test_data), unique = TRUE)
colnames(sentiment_notes) <- make.names(colnames(sentiment_notes), unique = TRUE)
```

```{r}
training_data <- training_data %>%
  left_join(sentiment_notes, by = "Match.ID.18Char")
```

### Compare Early vs. Late Stage Notes

```{r}
early_stage_threshold <- quantile(training_data$match_length, 0.25)
late_stage_threshold <- quantile(training_data$match_length, 0.75)
```

**Filter for early and late stages:**

```{r}
training_notes <- training_notes %>%
  left_join(training_data %>% 
              select(Match.ID.18Char, Match.Length), 
            by = c("Match ID 18Char" = "Match.ID.18Char"))
```

```{r}
early_stage_notes <- training_notes %>% filter(Match.Length <= early_stage_threshold)
late_stage_notes <- training_notes %>% filter(Match.Length >= late_stage_threshold)
```

**Compare word frequencies**

```{r}
early_word_freq <- early_stage_notes %>% count(word, sort = TRUE)
late_word_freq <- late_stage_notes %>% count(word, sort = TRUE)
```

**Plot top terms for comparison:**

```{r}
early_word_freq %>% top_n(20, n) %>% ggplot(aes(x = reorder(word, n), y = n)) +
  geom_col(fill = "skyblue") +
  coord_flip() +
  labs(title = "Top Words in Early Stage Notes")
```

```{r}
late_word_freq %>% top_n(20, n) %>% ggplot(aes(x = reorder(word, n), y = n)) +
  geom_col(fill = "coral") +
  coord_flip() +
  labs(title = "Top Words in Late Stage Notes")
```

### Predictive Model for Match Length

```{r}
model_data <- training_data %>%
  ungroup()%>%
  select(`Match.Length`, 
         days_between_calls, 
         positive, 
         negative) %>%
  drop_na()
```

```{r}
set.seed(123)
train_index <- createDataPartition(model_data$Match.Length, p = 0.8, list = FALSE)
train_set <- model_data[train_index, ]
test_set <- model_data[-train_index, ]
```

### Linear Regression:

```{r}
lm_model <- lm(Match.Length ~ ., data = train_set)

lm_predictions <- predict(lm_model, test_set)

lm_rmse <- sqrt(mean((lm_predictions - test_set$Match.Length)^2))
print(paste("Linear Regression RMSE:", lm_rmse))
```

RMSE of `15.89` means that, on average, the predicted **Match Length** is off by about **15.89 days** from the actual length.

### Random Forest:

```{r}
rf_model <- randomForest(Match.Length ~ ., data = train_set, ntree = 500)

rf_predictions <- predict(rf_model, test_set)

rf_rmse <- sqrt(mean((rf_predictions - test_set$Match.Length)^2))
print(paste("Random Forest RMSE:", rf_rmse))
```

#### **Random Forest Hyperparameters to Tune**

-   `ntree`: Number of trees in the forest.

-   `mtry`: Number of features to consider when looking for the best split.

-   `nodesize`: Minimum size of terminal nodes (affects model complexity).

-   `maxnodes`: Maximum number of terminal nodes.

```{r}
rf_grid <- expand.grid(
  mtry = c(2, 3, 4, 5, 6),
  ntree = c(100, 200, 300, 500),
  nodesize = c(1, 5, 10),
  maxnodes = c(10, 20, 30)
)

# Train the model using 5-fold cross-validation
set.seed(123)
rf_train_control <- trainControl(method = "cv", number = 5)
rf_model_tune <- train(
  Match.Length ~ .,
  data = train_set,
  method = "rf",
  trControl = rf_train_control,
  tuneGrid = rf_grid,
  importance = TRUE
)

# View the best parameters
print(rf_model_tune)
# Predict with the best model
rf_predictions <- predict(rf_model_tune, test_set)
rf_rmse <- sqrt(mean((rf_predictions - test_set$Match.Length)^2))
print(paste("Random Forest RMSE (Tuned):", rf_rmse))
```

### XGBoost:

```{r}
dtrain <- xgb.DMatrix(data = as.matrix(train_set %>% select(-Match.Length)), label = train_set$Match.Length)
dtest <- xgb.DMatrix(data = as.matrix(test_set %>% select(-Match.Length)), label = test_set$Match.Length)

params <- list(
  objective = "reg:squarederror",
  eta = 0.1,
  max_depth = 6,
  eval_metric = "rmse"
)

xgb_model <- xgboost(params = params, data = dtrain, nrounds = 100, verbose = 0)

xgb_predictions <- predict(xgb_model, dtest)

xgb_rmse <- sqrt(mean((xgb_predictions - test_set$Match.Length)^2))
print(paste("XGBoost RMSE:", xgb_rmse))

```

### Predict on Test-Truncated Data:

```{r}
test_data_prepared <- test_data %>%
  select(days_between_calls, positive, negative, neutral) %>%
  mutate_all(as.numeric) %>%
  replace_na(0)

# Make predictions on Test-Truncated using best model
test_dmatrix <- xgb.DMatrix(data = as.matrix(test_data_prepared))
test_predictions <- predict(xgb_model, test_dmatrix)

# Save predictions to CSV
submission <- data.frame(match_id = test_data$match_id, predicted_match_length = test_predictions)
```

```{r}
library(text2vec)
# Preprocess notes: tokenize, remove stopwords, etc.
notes_data <- train_set$Match.Support.Contact.Notes  # Assuming you have this column
corpus <- tolower(notes_data)
tokens <- word_tokenizer(corpus)
vectorizer <- vocab_vectorizer(vocabulary = create_vocabulary(itoken(tokens)))
dtm <- create_dtm(itoken(tokens), vectorizer)

# Create TF-IDF features
tfidf <- TfIdf$new()
dtm_tfidf <- tfidf$fit_transform(dtm)

# Combine TF-IDF features with other numeric features
model_data <- cbind(train_set, as.data.frame(dtm_tfidf))

# Train your model with the updated data
dtrain <- xgb.DMatrix(data = as.matrix(model_data %>% select(-Match.Length)), label = model_data$Match.Length)

```

# REDO

```{r warning=FALSE}
train_data_sent <- read_excel(training_file, sheet = 1)
test_data_sent <- read_excel(test_file, sheet = 1)
```

Checking column names

```{r}
colnames(train_data_sent)
colnames(test_data_sent)
```

```{r}
# Columns in train_data_sent but not in test_data_sent
missing_from_test <- setdiff(colnames(train_data_sent), colnames(test_data_sent))

# Columns in test_data_sent but not in train_data_sent
missing_from_train <- setdiff(colnames(test_data_sent), colnames(train_data_sent))

# Print results
cat("Columns in train_data_sent but missing from test_data_sent:\n")
print(missing_from_test)

cat("\nColumns in test_data_sent but missing from train_data_sent:\n")
print(missing_from_train)
```

### Sentiment Analysis Round 2

Function to compute sentiment score from text columns

```{r}
compute_sentiment <- function(text_column) {
  if (all(is.na(text_column))) return(rep(NA, length(text_column))) 
  sentiment_scores <- sentimentr::sentiment(text_column) %>%
    group_by(element_id) %>%
    summarize(sentiment = mean(sentiment, na.rm = TRUE))
  
  return(sentiment_scores$sentiment)
}
```

Compute sentiment scores from `Match Support Contact Notes`

```{r}
train_data_sent$Support_Notes_Sentiment <- compute_sentiment(train_data_sent$`Match Support Contact Notes`)
test_data_sent$Support_Notes_Sentiment <- compute_sentiment(test_data_sent$`Match Support Contact Notes`)
```

#### Compare Rationale for Match & Support Notes

Compute sentiment scores for Rationale for Match

```{r}
train_data_sent$Rationale_Sentiment <- compute_sentiment(train_data_sent$`Rationale for Match`)
test_data_sent$Rationale_Sentiment <- compute_sentiment(test_data_sent$`Rationale for Match`)
```

Compare Rationale and Support Notes Sentiment

```{r}
train_data_sent <- train_data_sent %>%
  mutate(Rationale_Match = ifelse((Support_Notes_Sentiment >= 0 & Rationale_Sentiment >= 0) |
                                  (Support_Notes_Sentiment < 0 & Rationale_Sentiment < 0), 
                                  "Aligned", "Misaligned"))

test_data_sent <- test_data_sent %>%
  mutate(Rationale_Match = ifelse((Support_Notes_Sentiment >= 0 & Rationale_Sentiment >= 0) |
                                  (Support_Notes_Sentiment < 0 & Rationale_Sentiment < 0), 
                                  "Aligned", "Misaligned"))
```

Visualize match alignment

```{r}
ggplot(train_data_sent, aes(x = Rationale_Match, fill = Rationale_Match)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Alignment of Rationale for Match with Support Notes Sentiment",
       x = "Sentiment Alignment", y = "Count")
```

```{r}
ggplot(train_data_sent, aes(x = Rationale_Match, fill = Rationale_Match)) +
  geom_bar() +
  scale_fill_manual(values = c("Aligned" = "#1f77b4", "Misaligned" = "#ff7f0e")) + # Custom colors
  theme_minimal(base_size = 14) + # Consistent base size for readability
  labs(
    title = "Alignment of Rationale for Match with Support Notes Sentiment",
    x = "Sentiment Alignment", 
    y = "Count"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), # Angle x-axis text
    legend.position = "none" # Remove legend if not needed
  )
```

```{r}
saveRDS(train_data_sent, "train_data_sent.rds")
saveRDS(test_data_sent, "test_data_sent.rds")
```

### Demographic Pattern Analysis

Function to check sentiment trends across different groups

```{r}
co
```

Analyze sentiment trends for different demographics

```{r}
age_analysis <- analyze_demographics(train_data_sent, "Big Age")
gender_analysis <- analyze_demographics(train_data_sent, "Big Gender")
race_analysis <- analyze_demographics(train_data_sent, "Big Race/Ethnicity")
education_analysis <- analyze_demographics(train_data_sent, "Big Level of Education")
```

Plot sentiment by demographics factor

```{r}
ggplot(age_analysis, aes(x = reorder(`Big Age`, avg_sentiment), y = avg_sentiment, fill = `Big Age`)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Average Sentiment Score by Big's Age",
       x = "Big's Age", y = "Average Sentiment")
```

```{r}
# Function for Sentiment Trend Analysis
ggplot(age_analysis, aes(
  x = factor(`Big Age`, levels = sort(unique(`Big Age`))),
  y = avg_sentiment,
  fill = `Big Age`
)) +
  geom_col() +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Average Sentiment Score by Big's Age",
    x = "Big's Age",
    y = "Average Sentiment",
    fill = "Age"
  ) +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, size = 16),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )



```

```{r}
ggplot(gender_analysis, aes(x = reorder(`Big Gender`, avg_sentiment), y = avg_sentiment, fill = `Big Gender`)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Average Sentiment Score by Big's Gender",
       x = "Big's Gender", y = "Average Sentiment")
```

```{r}
# Example for gender sentiment plot
ggplot(gender_analysis, aes(x = reorder(`Big Gender`, avg_sentiment), y = avg_sentiment, fill = `Big Gender`)) +
  geom_col()  + 
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Average Sentiment Score by Big's Gender",
    x = "Big's Gender", 
    y = "Average Sentiment"
  ) +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, size = 16),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

```

```{r}
ggplot(race_analysis, aes(x = reorder(`Big Race/Ethnicity`, avg_sentiment), y = avg_sentiment, fill = `Big Race/Ethnicity`)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Average Sentiment Score by Big's Race/Ethnicity",
       x = "Big's Race/Ethnicity", y = "Average Sentiment")
```

```{r}
ggplot(race_analysis, aes(x = reorder(`Big Race/Ethnicity`, avg_sentiment), y = avg_sentiment, fill = `Big Race/Ethnicity`)) +
  geom_col()  + 
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Average Sentiment Score by Big Race/Ethnicity",
    x = "Big Race/Ethnicity", 
    y = "Average Sentiment"
  ) +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, size = 16),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r}
ggplot(education_analysis, aes(x = reorder(`Big Level of Education`, avg_sentiment), y = avg_sentiment, fill = `Big Level of Education`)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Average Sentiment Score by Big's Level of Education",
       x = "Big's Level of Education", y = "Average Sentiment")
```

```{r}
ggplot(education_analysis, aes(x = reorder(`Big Level of Education`, avg_sentiment), y = avg_sentiment, fill = `Big Level of Education`)) +
  geom_col()  + 
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Average Sentiment Score by Big Level of Education",
    x = "Big Level of Education", 
    y = "Average Sentiment"
  ) +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, size = 16),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

### Alternative Match Evaluation

Checking whether demographic alignment affects sentiment

```{r}
train_data_sent <- train_data_sent %>%
  mutate(Same_Gender_Match = ifelse(`Big Gender` == `Little Gender`, "Same Gender", "Different Gender"),
         Same_Race_Match = ifelse(`Big Race/Ethnicity` == `Little Participant: Race/Ethnicity`, "Same Race", "Different Race"))
```

Sentiment analysis based on match characteristics

```{r}
gender_match_analysis <- train_data_sent %>%
  group_by(Same_Gender_Match) %>%
  summarize(avg_sentiment = mean(Support_Notes_Sentiment, na.rm = TRUE))

race_match_analysis <- train_data_sent %>%
  group_by(Same_Race_Match) %>%
  summarize(avg_sentiment = mean(Support_Notes_Sentiment, na.rm = TRUE))
```

Visualize gender-based matching impact

```{r}
ggplot(gender_match_analysis, aes(x = Same_Gender_Match, y = avg_sentiment, fill = Same_Gender_Match)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Impact of Gender Matching on Sentiment",
       x = "Match Type", y = "Average Sentiment")
```

```{r}
# Improved gender-based sentiment match plot
ggplot(gender_match_analysis, aes(x = Same_Gender_Match, y = avg_sentiment, fill = Same_Gender_Match)) +
  geom_col() +
  scale_fill_manual(values = c("Same Gender" = "#2ca02c", "Different Gender" = "#d62728")) + # Custom colors
  theme_minimal(base_size = 14) +
  labs(
    title = "Impact of Gender Matching on Sentiment",
    x = "Match Type", 
    y = "Average Sentiment"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), # Angle x-axis text for readability
    plot.title = element_text(hjust = 0.5, size = 16),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

```

Saving Data for future use

```{r}
write.csv(train_data, "data/Processed_Train_Data.csv", row.names = FALSE)
write.csv(test_data, "data/Processed_Test_Data.csv", row.names = FALSE)
```

### Correlations

```{r}
cor(train_data_sent$Support_Notes_Sentiment, train_data_sent$`Match Length`, use = "complete.obs")
```

```{r}
table(train_data_sent$Rationale_Match, train_data_sent$`Closure Reason`)
```

```{r}
table(train_data_sent$Same_Gender_Match, train_data_sent$Rationale_Match)
table(train_data_sent$Same_Race_Match, train_data_sent$Rationale_Match)
```

model

```{r}
model <- glm(`Closure Reason` ~ Support_Notes_Sentiment + Rationale_Match + Same_Gender_Match + Same_Race_Match, 
             data = train_data_sent, family = binomial)

summary(model)
```

```{r}
train_data$Match_Success <- ifelse(
  train_data$`Stage` == "Active" | 
  train_data$`Match Length` >= 24 |  
  grepl("Aged Out|Successful Match Closure", train_data$`Closure Reason`, ignore.case = TRUE), 
  1, 0
)

table(train_data$Match_Success) 

```

```{r}
model <- glm(Match_Success ~ Support_Notes_Sentiment + Rationale_Match + Same_Gender_Match + Same_Race_Match, 
             data = train_data_sent, family = binomial)
summary(model)

```

```{r}
cor(train_data_sent$Match_Success, train_data_sent$`Match Length`, use = "complete.obs")

```

```{r}
ggplot(train_data_sent, aes(x = factor(Match_Success), fill = factor(Same_Gender_Match))) +
    geom_bar(position = "fill") +
    theme_minimal() +
    labs(title = "Match Success by Gender", x = "Match Success", y = "Proportion")

```

```{r}
sapply(train_data_sent[, c("Rationale_Match", "Same_Gender_Match", "Same_Race_Match")], function(x) length(unique(x)))

```

```{r}
train_data_sent$Rationale_Match <- as.factor(train_data_sent$Rationale_Match)
train_data_sent$Same_Gender_Match <- as.factor(train_data_sent$Same_Gender_Match)
train_data_sent$Same_Race_Match <- as.factor(train_data_sent$Same_Race_Match)

```

```{r}
table(train_data_sent$Same_Gender_Match)
unique(train_data_sent$Same_Gender_Match)

```

Convert `Same_Gender_Match` to a Proper Factor

```{r}
train_data_sent$Same_Gender_Match <- factor(train_data_sent$Same_Gender_Match, 
                                       levels = c("Different Gender", "Same Gender"))
```

```{r}
# Impute NA with "Same Gender" (since it’s the most frequent)
train_data_sent$Same_Gender_Match[is.na(train_data_sent$Same_Gender_Match)] <- "Same Gender"
train_data_sent$Same_Gender_Match <- as.factor(train_data_sent$Same_Gender_Match)

```

```{r}
model <- glm(Match_Success ~ Support_Notes_Sentiment + Rationale_Match + Same_Gender_Match + Same_Race_Match, 
             data = train_data_sent, family = binomial)

summary(model)
```

```{r}
sapply(train_data_sent[, c("Rationale_Match", "Same_Gender_Match", "Same_Race_Match")], function(x) unique(x))

```

```{r}
table(train_data_sent$Rationale_Match)
table(train_data_sent$Same_Gender_Match)
table(train_data_sent$Same_Race_Match)

```

Investigating `Same_Race_Match`

```{r}
sum(is.na(train_data_sent$Same_Race_Match))

```

```{r}
table(is.na(train_data_sent$Same_Race_Match), train_data_sent$Match_Success)

```

```{r}
model <- lm(Match_Length ~ Support_Notes_Sentiment + Rationale_Match + Same_Gender_Match, 
            data = train_data_sent)
summary(model)

```

```{r warning=FALSE}
match_length_train_data <- read_excel(training_file, sheet = 1)
match_length_test_data <- read_excel(test_file, sheet = 1)
```

```{r}
summary(match_length_train_data$`Match Length`)  # Check min, max, median, mean
hist(match_length_train_data$`Match Length`, breaks=30, main="Distribution of Match Length", xlab="Months")
sum(is.na(match_length_train_data$`Match Length`))  # Count NAs

```

```{r}
# Select some potential predictors
potential_predictors <- c(
  "Big Gender", "Little Gender", "Big Race/Ethnicity", "Little Participant: Race/Ethnicity",
  "Rationale for Match", "Big Car Access", "Big Days Interview to Match", 
  "Big Level of Education", "Program", "Program Type"
)

# Check structure and unique values
sapply(match_length_train_data[, potential_predictors], function(x) {
  c(Num_NA = sum(is.na(x)), Unique_Values = length(unique(x)))
})

```

```{r}
# Basic linear model with selected predictors
lm_model <- lm(`Match Length` ~ `Big Gender` + `Little Gender` + `Program Type` + 
               `Big Race/Ethnicity` + `Little Participant: Race/Ethnicity`, 
               data = match_length_train_data)

# Summary of the model
summary(lm_model)

```

```{r}
coef_df <- data.frame(
  Feature = names(coef(lm_model)),
  Coefficient = coef(lm_model)
)

ggplot(coef_df, aes(x = reorder(Feature, Coefficient), y = Coefficient)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  coord_flip() +
  labs(title = "Linear Model Coefficients", x = "Feature", y = "Coefficient")

```

```{r}
# Calculate residuals
residuals <- residuals(lm_model)

# Create residual plot
ggplot(data.frame(Fitted = fitted(lm_model), Residuals = residuals), aes(x = Fitted, y = Residuals)) +
  geom_point(color = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residual Plot", x = "Fitted Values", y = "Residuals")

```

```{r}
# Assuming you have a dataset with actual values corresponding to your model
# Replace 'actual_values' with the actual response variable from your dataset
model_levels <- lm_model$xlevels$`Little Participant: Race/Ethnicity`

# Filter the train data to only rows with factor levels used in the model
match_length_train_data_fixed <- match_length_train_data[
  match_length_train_data$`Little Participant: Race/Ethnicity` %in% model_levels, 
]

# Now predict using the cleaned dataset
train_predictions <- predict(lm_model, newdata = match_length_train_data_fixed)


# Add the actual values from training data
actual_train_values <- match_length_train_data_fixed$`Match Length`

# Now the plot will work
ggplot(data.frame(Actual = actual_train_values, Predicted = train_predictions), aes(x = Actual, y = Predicted)) +
  geom_point(color = "darkgreen", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Predicted vs Actual (Training Data)",
    x = "Actual Match Length",
    y = "Predicted Match Length"
  ) +
  theme_minimal()

```

```{r}
# Make sure test data uses same levels
match_length_test_data_fixed <- match_length_test_data[
  match_length_test_data$`Little Participant: Race/Ethnicity` %in% training_levels, 
]

# Predict on test data
test_predictions <- predict(lm_model, newdata = match_length_test_data_fixed)

# Get actuals from filtered test data
actual_test_values <- match_length_test_data_fixed$`Match Length`

# Calculate residuals
residuals_test <- actual_test_values - test_predictions

# RMSE and R-squared
rmse_test <- rmse(actual_test_values, test_predictions)
sse_test <- sum(residuals_test^2)
sst_test <- sum((actual_test_values - mean(actual_test_values))^2)
r_squared_test <- 1 - sse_test/sst_test

# Print results
cat("Test RMSE:", round(rmse_test, 2), "\n")
cat("Test R²:", round(r_squared_test, 4), "\n")
```

```{r}
# Store levels from training data
race_levels <- levels(match_length_train_data$`Little Participant: Race/Ethnicity`)
gender_levels <- levels(match_length_train_data$`Little Gender`)

# Keep only rows where both Race and Gender levels are in the training set
match_length_test_data_fixed <- match_length_test_data[
  match_length_test_data$`Little Participant: Race/Ethnicity` %in% race_levels &
  match_length_test_data$`Little Gender` %in% gender_levels,
]

# Predict on filtered test data
test_predictions <- predict(lm_model, newdata = match_length_test_data_fixed)

# Actual values
actual_test_values <- match_length_test_data_fixed$`Match Length`

# Residuals
residuals_test <- actual_test_values - test_predictions

# RMSE and R²
library(Metrics)
rmse_test <- rmse(actual_test_values, test_predictions)
sse_test <- sum(residuals_test^2)
sst_test <- sum((actual_test_values - mean(actual_test_values))^2)
r_squared_test <- 1 - sse_test / sst_test

# Output metrics
cat("Test RMSE:", round(rmse_test, 2), "\n")
cat("Test R²:", round(r_squared_test, 4), "\n")

```

```{r}
match_length_test_data_fixed <- match_length_test_data[
  match_length_test_data$`Little Participant: Race/Ethnicity` %in% levels(match_length_train_data$`Little Participant: Race/Ethnicity`) &
  match_length_test_data$`Little Gender` %in% levels(match_length_train_data$`Little Gender`),
]

# Predict on cleaned test data
test_predictions <- predict(lm_model, newdata = match_length_test_data_fixed)

output_df <- match_length_test_data_fixed
output_df$Predicted_Match_Length <- test_predictions

# Export to CSV
write.csv(data.frame(Predicted_Match_Length = test_predictions), "match_length_predictions.csv", row.names = FALSE)

```

```{r}
glimpse(match_length_train_data)

# Or view the first few rows
head(match_length_train_data)
```

```{r}
# Check for missing values in the training data
sum(is.na(match_length_train_data))

```

```{r}
target_variable <- "Match Length"

# All columns except the target
features <- setdiff(names(match_length_train_data), target_variable)
```

```{r}
set.seed(1501) # Set seed for reproducibility
train_index <- createDataPartition(match_length_train_data[[target_variable]], p = 0.8, list = FALSE)
train_data <- match_length_train_data[train_index, ]
test_data <- match_length_train_data[-train_index, ]
```

```{r}
# Preprocess the data
preProc <- preProcess(train_data[, features], method = c("center", "scale"))

# Apply preprocessing to the training and test data
train_data_processed <- predict(preProc, train_data[, features])
test_data_processed <- predict(preProc, test_data[, features])

```

```{r}
formula <- as.formula(paste("`", target_variable, "` ~ ", 
                            paste(sapply(features, function(x) paste("`", x, "`", sep = "")), collapse = " + "), sep = ""))

# Fit the linear model
model <- lm(formula, data = match_length_train_data)
# View model summary
summary(model)
```

```{r}
features_cleaned <- features[sapply(match_length_train_data[, features], function(x) length(unique(x)) > 1)]

# Optionally, remove features with very few unique values (e.g., less than 5 unique values)
features_cleaned <- features_cleaned[sapply(match_length_train_data[, features_cleaned], function(x) length(unique(x)) > 4)]

# Ensure that categorical variables are treated as factors
match_length_train_data[features_cleaned] <- lapply(match_length_train_data[features_cleaned], function(x) {
  if (is.character(x)) {
    factor(x)
  } else {
    x
  }
})

# Create the formula again with the cleaned features
formula <- as.formula(paste("`", target_variable, "` ~ ", 
                            paste(sapply(features_cleaned, function(x) paste("`", x, "`", sep = "")), collapse = " + "), sep = ""))

# Fit the model
model <- lm(formula, data = match_length_train_data)
```
